---
title: "Influence of Trophic Position and Body Size on Shark Metabolic Rate"
date: "2024-12-06"
output:
  html_document:
    toc: yes
    toc_float: yes
    collapsed: no
    smooth_scroll: yes
---
Group 1: Emma Booth (1149027), Annika Laurin (1183145), Samantha Mok (1126756), Kayla Perez (1175585), Farah Sadoon (1302190)

## Introduction

## Methods
- Body mass alone PGLS
- Trophic position alone PGLS
- Both together PGLS
- AIC model on all 3 
- Build new folder in GitHub without unnecessary steps that can be run through all at once
- Use trait mapping to solidify ideas found in AIC

insert cleaning data code
```{r}
#load libraries
library(readxl)
library(readr)
library(janitor)
library(dplyr)
library(stringr)
library(ape)
library(ggplot2)
library(phylolm)
library(phytools)
library(tidyverse)
library(MuMIn)
library(viridis)
library(gridExtra)



-------------------------------------------------------------------------------------
### CLEANING ###


FULL_Rmax_LHT_MR_ms <- read_csv("00_raw/00b_FULL_Rmax_LHT_MR_ms.csv")

# find column names
str(FULL_Rmax_LHT_MR_ms)
column_names <- names(FULL_Rmax_LHT_MR_ms)

# clean datasets
main_dataset1 <- FULL_Rmax_LHT_MR_ms %>%
  select(-TeleostOrElasmo, -MROriginLocation, -MRCitation, -DFtype )

# Convert column names to snake case
colnames(main_dataset1) <- make_clean_names(names(main_dataset1), case = "snake")

# removing upper case letters and adding "_"
main_dataset2 <- main_dataset1 %>%
  clean_names() %>%                # Clean column names
  mutate(across(everything(), ~ str_replace_all(., " ", "_"))) %>%  # Replace spaces with underscores
  mutate(across(everything(), tolower)) %>%  # Convert all to lowercase
  mutate(scientific_name = str_to_title(scientific_name))  # Convert scientific_name to title case

### Fishbase trophic position data:

trophic_position_clean <- read_csv("01_processed_datasets/01c_trophic_position_clean.csv")

## adding underscores to match main_dataset2
trophic_position_clean <- trophic_position_clean %>%
  mutate(scientific_name = gsub(" ", "_", scientific_name))


main_dataset3 <- left_join(main_dataset2, trophic_position_clean, by = "scientific_name")
str(main_dataset3)

main_dataset3$mr_mass_grams <- as.numeric(as.character(main_dataset3$mr_mass_grams))

```

insert tree making
```{r}

```

PGLS and OLS
```{r}

```

## Results
- Fix line of best fit on graphs
- Model comparisons

PGLS and OLS results
```{r}

```

Trait Mapping
```{r}

```

## Discussion

## References


