---
title: "Influence of Trophic Position and Body Size on Shark Metabolic Rate"
date: "2024-12-06"
output:
  html_document:
    toc: yes
    toc_float: yes
    collapsed: no
    smooth_scroll: yes
---
Group 1: Emma Booth (1149027), Annika Laurin (1183145), Samantha Mok (1126756), Kayla Perez (1175585), Farah Sadoon (1302190)

##Load Libraries

```{r}
library(readxl)
library(readr)
library(janitor)
library(dplyr)
library(stringr)
library(ape)
library(ggplot2)
library(phylolm)
library(phytools)
library(tidyverse)
library(MuMIn)
library(viridis)
library(gridExtra)
```

## Introduction

## Methods
### 2.1 Data preparation 
	The necessary libraries for data manipulation, analysis, and visualization were loaded. The raw dataset, FULL_Rmax_LHT_MR_ms.csv, was imported, and unnecessary columns, including TeleostOrElasmo, MROriginLocation, MRCitation, and DFtype, were removed. Column names were standardized to snake case. The cleaned dataset was merged with trophic position data from the online database FishBase using the scientific_name column. Rows with missing values (NA) were removed, and numerical columns were converted to their appropriate data types for further analysis.
### 2.2 Transformations
	Performed Shapiro-Wilk tests to assess the normality of columns containing metabolic rate, body mass, and trophic position data. To achieve normality, applied fourth-root transformations to the metabolic rate and body mass variables, followed by standardizing the transformed data using z-scores. Created histograms of metabolic rate, body mass, and trophic position for both raw and transformed data to visually evaluate distributions. Identified outliers in the standardized fourth-root-transformed variables using the interquartile range (IQR) method.
### 2.3 Ordinary Least Squares (OLS) Modelling
	A simple linear regression was performed to examine the relationships between standardized whole-organism metabolic rate and standardized trophic position, standardized whole-organism metabolic rate and standardized body mass, and standardized trophic position and standardized body mass. A multiple linear regression was conducted to evaluate the combined effects of standardized trophic position and standardized body mass on standardized whole-organism metabolic rate. Residuals from the multiple regression model were extracted for analysis and visualization. Scatter plots of residuals against standardized trophic position and standardized body mass were created to identify potential patterns and deviations. A diagnostic plot of fitted values versus residuals was generated to assess model assumptions and overall fit.
### 2.4 Phylogenetic Tree
  Loaded the full teleost chronogram and the chondro subset phylogenetic tree from the Gravel et al., 2024 dataset. Species names in the phylogenetic tree were updated to correct the nomenclature and match the dataset created in section 2.2. Species names were extracted from both the dataset and the phylogenetic tree, and a comparison was made to identify any mismatches between the two datasets. The phylogenetic tree was then pruned to include only the species present in the dataset.
### 2.5 Phylogenetic Least Squares (PGLS) Modelling
Reordered the dataset to align with the order of tree tips by filtering for species present in both the tree and dataset, then arranging the rows according to the sequence of tree tips. Verified the root edge length of the pruned tree and corrected it by setting the value to 0. Fitted three PGLS models: one predicting whole-organism metabolic rate using both trophic position and body mass, another using only trophic position, and a third using only body mass. Performed a model comparison to identify the best models based on delta AIC values.
### 2.6 Trait Mapping
Calculated the phylogenetic signal for average whole-organism metabolic rate using the lambda method. Grouped the dataset by species and calculated the mean values for standardized whole-organism metabolic rate, trophic position, and body mass. Created 3 phylogenetic trees with the traits metabolic rate, trophic position, and body mass respectively. 


insert cleaning data code
```{r}
FULL_Rmax_LHT_MR_ms <- read_csv("00_raw/00b_FULL_Rmax_LHT_MR_ms.csv")

# find column names
str(FULL_Rmax_LHT_MR_ms)
column_names <- names(FULL_Rmax_LHT_MR_ms)

# clean datasets
main_dataset1 <- FULL_Rmax_LHT_MR_ms %>%
  select(-TeleostOrElasmo, -MROriginLocation, -MRCitation, -DFtype )

# Convert column names to snake case
colnames(main_dataset1) <- make_clean_names(names(main_dataset1), case = "snake")

# removing upper case letters and adding "_"
main_dataset2 <- main_dataset1 %>%
  clean_names() %>%                # Clean column names
  mutate(across(everything(), ~ str_replace_all(., " ", "_"))) %>%  # Replace spaces with underscores
  mutate(across(everything(), tolower)) %>%  # Convert all to lowercase
  mutate(scientific_name = str_to_title(scientific_name))  # Convert scientific_name to title case

### Fishbase trophic position data:

trophic_position_clean <- read_csv("01_processed_datasets/01c_trophic_position_clean.csv")

## adding underscores to match main_dataset2
trophic_position_clean <- trophic_position_clean %>%
  mutate(scientific_name = gsub(" ", "_", scientific_name))


main_dataset3 <- left_join(main_dataset2, trophic_position_clean, by = "scientific_name")
str(main_dataset3)

main_dataset3$mr_mass_grams <- as.numeric(as.character(main_dataset3$mr_mass_grams))

```

insert tree making
```{r}

```

PGLS and OLS
```{r}

```

## Results

### 3.1 OLS
  Our first linear regression model examined the relationship between metabolic rate and trophic position, which revealed a significant positive relationship (slope = 0.638, p-value < 2 x 10-16), explaining 40.7% of the variance in metabolic rate (R2 = 0.407) with a residual standard error (RSE) of 0.7703. This indicates that when considered as the only explanatory variable, trophic position is a strong predictor for metabolic rate. 
  
  In comparison, the results from our metabolic rate vs. body mass showed a significantly stronger positive relationship (slope = 0.769, p-value < 2 x 10-16) that explained a higher proportion of variance in metabolic rate (R2= 0.592) with a lower RSE of 0.6392. Therefore, these results propose that the body mass model is a better fit when compared to the trophic position-only model.
  
  When both variables were included in a multiple linear regression model, body mass remained the dominant predictor (slope = 0.727, p-value < 2 x 10-16), while the trophic position was weak and non-significant (slope = 0.053, p = 0.072). This model explained 59.3% of variance in metabolic rate (R2 = 0.593) with an RSE of 0.6386, which only slightly differed from the body mass-only model. These findings suggest that trophic position has minimal influence when body mass is accounted for in the model. The residuals of the combined model were also evenly distributed around zero (702 positive, 629 negative values) with no patterns, confirming the validity of the model. 

### 3.2 PGLS
  In order to account for phylogenetic relationships among species, we utilized a PGLS model to assess the relationship between variables metabolic rate, trophic position, and body mass. In the combined model (trophic position + body mass) body mass remained a significant predictor (β = 1.253, p-value < 0.001), while trophic position did not show a significant effect (β = 0.0036, p-value = 0.9752). The model explained 79.5% of the variance in metabolic rate and had a strong phylogenetic signal of 0.814 meaning that metabolic rates were highly conserved in closely related species. 

  The body mass-only model showed to have the lowest AIC value (AIC = 73.03) and explained a similar proportion of variance as the combined model. Body mass persisted as a strong predictor for metabolic rate; however, the trophic position-only model showed a much poorer fit with a high AIC value (AIC = 113.58) and a non-significant effect (β = 0.604, p = 0.9752). 

Our model selection then identified the body mass-only model as the best-supported model (ΔAIC = 0), which accounted for 80.2% of the model weight. The combined model was the second best model (ΔAIC = 2.79, model weight = 19.8%) and trophic position performed the worst (ΔAIC = 40.55). In the model comparison plot, the body mass-only model was the only one to fall into the ΔAIC < 2 threshold (fig 4) indicating the dominant role of body mass in explaining metabolic rate variability across shark species. 

PGLS and OLS results
```{r}

```


### 3.3 Trait Mapping
	Trait mapping revealed no clear pattern in the distribution of metabolic rates associated with ecological roles across species since some carnivorous and omnivorous species both had high metabolic rates, therefore our ecological hypothesis is not supported. In contrast, body mass showed some correlation to metabolic rate since larger species like Oncorhynchus nerka exhibited higher metabolic rates. However, there are some exceptions such as small species like Gadus morhua having moderate rates indicating that other factors may play a role in determining metabolic rates. 
	
  Our results strongly support the “Body Size” hypothesis demonstrating that body mass is the primary driver of metabolic rate across shark species, outperforming trophic position in explaining variation. While trophic position had some predictive value when considered alone, its influence was minimal when body mass was included. 

Trait Mapping
```{r}

```

## Discussion
The relationship between body size and MR is a key component of the metabolic theory (Brown et al., 2004). Our research suggests that body size can be a strong predictor of MR in individuals, enhancing our understanding of marine metabolism (Brown et al., 2004). Typically, MR is measured in controlled laboratory settings where basal metabolic rate (BMR) is assessed when the organism is inactive (Brown et al., 2004). However, in natural environments, field metabolic rates (FMR) are typically higher than BMR, as organisms must allocate energy for activities like foraging, predator avoidance, physiological regulation, and other maintenance processes (Brown et al., 2004). This discrepancy between BMR and FMR is an important area of study, as it provides insights into energy expenditure, survival strategies, and ecological adaptations.
  
By understanding the relationship between trophic position and metabolic rates, we can better identify species that are sensitive to overfishing (Gravel et al., 2024). Since sharks often exhibit slow life histories characterized by slow growth, they are more sensitive to overfishing (Gravel et al., 2024). By better understanding the relationship between trophic position, metabolisms and body size, our research can help identify where different species fall along a spectrum of sensitivity to population declines.
  
By examining how metabolism and life history traits, such as body size and trophic position, are connected, we can gain a deeper understanding of how species evolve in response to their environment (Gravel et al., 2024). This approach can also enhance our understanding of Elasmobranch evolution and shed light on how these species have survived for so long.
  
Key questions emerging from our research are focused on understanding how climate change may interact with marine populations. With rising global ocean temperatures, understanding how Elasmobranch metabolism responds to environmental changes can help predict shifts in their populations and ranges (Osgood et al., 2021). 

## References


